---
title: "R formation"
author: "Elodie Darbo / Cyril Dourthe"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Discovering the R console

R is an interpreted language. If you write a code line in the console and validate it, you will instant see the result of your code.

R can do everything !

```{r simple code}
42
```

If you just type a number, R will interpret the simplest way : by returning the number you entered.

```{r addition}
2+3
```

Unlike the previous line, R will not return "2+3", but it interpret the "+" symbol as an addition.

```{r multiplication}
4*5
```

You can do the same manipulation with a multiplication, the "\*" symbol is interpreted as a multiplication

```{r division}
6/4
```

Same phenomenon, but with the division, the "/" symbol is interpreted as a division

```{r sequence1}
1:10
```

This line is different. The ":" is called colon. With this symbol you can generate a sequence of number.

```{r sequence2}
8:-9
```

Same as the previous line, here we generate a sequence of number with negative and positive numbers.

```{r error}
1,2
```

```{r decimal}
1.2
```

R is demanding when it comes to the use of certain symbols, particularly decimal numbers. In french, we used to write the decimal with a comma like in this example. But in english, the decimal separator is a point not a comma.
If you write it with a comma, an error will occured, R can't interpret what you wrote.

```{r comment}
# This is a comment
```

You can write entire lines of text. To do this, simply use the "#" symbol at the beginning of the line.
This code line is called a comment, and it is used to indicate that the different lines of code we write do.


## Work with variables


