---
title: "ggplot2 tutorial"
format:
  html:
    code-fold: true
editor: visual
---

```{=html}
<style>
  details > summary {  
    padding: 4px;  
    width: 400px;  
    background-color: #eeeeee;  
    border: none;  
    box-shadow: 1px 1px 2px #bbbbbb;  
    cursor: pointer;
    
  }
  details > p {  
    background-color: #E0F8EC;  
    padding: 4px;  
    margin: 0;  
    box-shadow: 1px 1px 2px #bbbbbb;
  }
</style>
```
```{r, setup, echo=F, warning=FALSE,message=FALSE}
require("knitr")
library(ggplot2)
library(ggpubr)
library(corrplot)
library(pheatmap)
```

## Introduction

Describe the study and the data ... to be done

## Load the data

#### You first need to load the data.

Follow the procedure using the interface to load the file or try to use the function `read.table.`

<!--# Describe rows and columns -->

Clinical data: `TCGA_LUAD_subset.tsv`, name the variable `clinical`

Survival data: `TCGA_LUAD_subset_survival.tsv`, name the variable `surv_dat`

Expression data : `TCGA_LUAD_expression.tsv,` name the variable `luad_expr`

```{r}
# or you can try to adapt this code
clinical <- read.table("data/TCGA_LUAD_subset.tsv",sep="\t",head=T)
head(clinical)
surv_dat <- read.table("data/TCGA_LUAD_subset_survival.tsv",sep="\t",head=T)
head(surv_dat)
luad_expr <- read.table("data/TCGA_LUAD_expression.tsv",sep="\t",head=T,row=1)
luad_expr[1:6,1:6]

```

#### Install ggplot2 and load the library

```{r, eval=FALSE}
install.packages("ggplot2")
library(ggplot2)
```

## Display your first plot step by step

Before unfolding the code and answers boxes, try to find the solution from what we have seen in the lecture or by **searching on the web**, it should become an automatism, no one knows everything !

1.  Use the function `ggplot` alone.

```{r}
ggplot()
```

A default grey background is displayed.

2.  Display the axis: Add the data and choose the x and y axis from the column names that you can display using `colnames`. For instance, we want to evaluate the distribution of the MAPK activity signature (y-axis) in function of the known oncogenes (x-axis).

```{r}
# Display the column names
colnames(clinical)
ggplot(data = clinical, aes(x=known.oncogenes, y=signature_MAPK))
```

<details>

<summary>Answer</summary>

The default theme displays a grey background and a white grid. We will see that is easy to change this features.

Note that the axis are automatically labelled with the column names.

</details>

3.  Display the geometric shape: **boxplot**

**Add a layer** to the plot. Which is the function to use ?

<details>

<summary>Key words to write on your favorite browser</summary>

-   `R ggplot2 boxplot`

</details>

```{r}
ggplot(data = clinical, aes(x=known.oncogenes, y=signature_MAPK)) +
  geom_boxplot()
```

## Customize by adding layers and modifying aesthetics

1.  Change the x-axis labels: we want to make it prettier: e.g Known Oncogenes and the y-axis to make it more explicit: e.g MAPK activity signature

```{r}
ggplot(data = clinical, aes(x=known.oncogenes, y=signature_MAPK)) +
  geom_boxplot() + 
  labs(x="Known Oncogenes", y="MAPK activity signature")
```

2.  Change the global theme to have a white background more suitable for figures and projection

<details>

<summary>Key words to write on your favorite browser</summary>

-   `R ggplot2 theme white background`

</details>

```{r}
ggplot(data = clinical, aes(x=known.oncogenes, y=signature_MAPK)) +
  geom_boxplot() + 
  labs(x="Known Oncogenes", y="MAPK activity signature") +
  theme_bw()
```

3.  Change the color and fill of the boxplots

-   All the boxplots in red and filled with grey. You can search for how colors are named in R if you want to try your favorite colors. Many palettes are available through dedicated packages. (My favorite function to call for colors is `colours()` systematically installed and loaded with R, you can try some of my preferred ones: `colours()[124]`, `colours()[613]`, `colours()[53]`, `colours()[144]`).

<details>

<summary>Key words to write on your favorite browser</summary>

-   `R ggplot2 colors`

```{r}
# the fonction colours() returns the name of the color
colours()[124]
# use the function c() to concat several elements, we call it a vector
c(colours()[613], colours()[53], colours()[144]) 
```

-   `R ggplot2 boxplot color`

</details>

```{r}
ggplot(data = clinical, aes(x=known.oncogenes, y=signature_MAPK)) +
  geom_boxplot(color="red", fill="grey") +
  labs(x="Known Oncogenes", y="MAPK activity signature") +
  theme_bw()
```

You can play globally on the color of the borders (parameter `color`), the filling (parameter `fill`), the transparency (parameter `alpha`)) and more. The accessible aesthetics can vary according to the geometric shape, form. instance for lines you will have the width and the type.

4.  Change the color of the boxplots in function of the known oncogenes

-   During the previous step the same colors have been applied to all boxplots, the goal here is to use the information contained in a column of our dataset (`known.oncogenes`). This is part of the aesthetics. Based on the current code, how would you add this information (2 possibilities) ?

***Option 1***

```{r, eval=F}
# Directly in the aesthetics of the ggplot function that will be accessible for all
# layers
ggplot(data = clinical, aes(x=known.oncogenes, y=signature_MAPK, color=known.oncogenes)) + # Add the new aesthetics color
  geom_boxplot() + # remove the global colors
  labs(x="Known Oncogenes", y="MAPK activity signature") +
  theme_bw()
```

***Option 2***

```{r}
# Within the geom_boxplot() function which will define the aesthetics 
# with aes() only for this layer
ggplot(data = clinical, aes(x=known.oncogenes, y=signature_MAPK)) +
  geom_boxplot(aes(color=known.oncogenes)) + 
  labs(x="Known Oncogenes", y="MAPK activity signature") +
  theme_bw()
```

-   By default the colors are not very contrasted ... you can customize them by i) defining your own palette optionally by **attributing a color to each category** ; ii) find a predefined palette to help you to define the colors. To do this task, you will need to add a new layer by using a function that help in **scaling the colors manually**.

i\) To define your own palette, you need to concatenate as many colors as the number of categories, here 12 colors. You can assign the colors to the categories by building a **named vector**.

```{r}
# We use the function c(), to name the elements of a vector you use the name you want
# and you assign the values with the sign =
palette.known.onco <- c(ALK="darkgrey",
                        BRAF_act=colours()[613],
                        BRAF.non=colours()[11],
                        EGFR=colours()[128],
                        ERBB2=colours()[76],
                        MAP2K1=colours()[509],
                        MET=colours()[121],
                        NF1=colours()[468],
                        None=colours()[92],
                        RIT1=colours()[642],
                        ROS1=colours()[34],
                        KRAS="black")
# Display the content
palette.known.onco
```

Now you have your colors, add the layer to the ggplot object and use the variable `palette.known.onco` you just created.

<details>

<summary>Key words to write on your favorite browser</summary>

-   R ggplot2 scaling color manually

</details>

```{r}
ggplot(data = clinical, aes(x=known.oncogenes, y=signature_MAPK)) +
  geom_boxplot(aes(color=known.oncogenes)) + 
  labs(x="Known Oncogenes", y="MAPK activity signature") +
  theme_bw() +
  scale_colour_manual(values=palette.known.onco)
```

ii\) R natively provides a few continuous color palettes which means that you do not need to install and load any library. Also when you install `ggplot2`, it installs some dependencies (packages that `ggplot2` needs to work properly), among them it installs the packages `RColorBrewer`, `viridis`, `paletteer`, `khroma` etc. and have implemented specific functions to use the palettes, it means that those packages are installed but not loaded. You can see which functions are available when typing `scale_color_` on RStudio. If you want to use the functions from these packages you need to load the package but it is not mandatory. Finally, a lot more palettes are available through packages.

***Native palettes***

```{r}
# rainbow, heat.colors, terrain.colors, topo.colors, cm.colors
ggplot(data = clinical, aes(x=known.oncogenes, y=signature_MAPK)) +
  geom_boxplot(aes(color=known.oncogenes)) + 
  labs(x="Known Oncogenes", y="MAPK activity signature") +
  theme_bw() +
  scale_colour_manual(values=terrain.colors(12))
```

***Viridis palettes***

```{r}
# options: viridis, magma, plasma, inferno, cividis
ggplot(data = clinical, aes(x=known.oncogenes, y=signature_MAPK)) +
  geom_boxplot(aes(color=known.oncogenes)) + 
  labs(x="Known Oncogenes", y="MAPK activity signature") +
  theme_bw() +
  scale_colour_viridis_d(option="inferno")
```

5.  Customize **theme** elements by **rotating** the **x-axis labels** by 45°

<details>

<summary>Key words to write on your favorite browser</summary>

-   `R ggplot2 rotate x axis labels`

</details>

```{r}
ggplot(data = clinical, aes(x=known.oncogenes, y=signature_MAPK)) +
  geom_boxplot(aes(color=known.oncogenes)) + 
  labs(x="Known Oncogenes", y="MAPK activity signature") +
  theme_bw() +
  scale_colour_viridis_d(option="inferno") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
```

6.  Flip the plot to get horizontal boxplots

There is two ways, the first one is very simple and consists in inverting x and y aesthetics, the second one involves the **addition** of a new layer to **flip the axis** (preferred option and it provides more flexibility). Lets keep our own color palette `palette.known.onco`

***Option 1***

```{r, eval=F}
ggplot(data = clinical, aes(y=known.oncogenes, x=signature_MAPK)) +
  geom_boxplot(aes(color=known.oncogenes)) + 
  labs(x="Known Oncogenes", y="MAPK activity signature") +
  theme_bw() +
  scale_colour_manual(values=palette.known.onco)
```

***Option 2***

<details>

<summary>Key words to write on your favorite browser</summary>

-   `R ggplot2 flip axis`

</details>

```{r}
ggplot(data = clinical, aes(x=known.oncogenes, y=signature_MAPK)) +
  geom_boxplot(aes(color=known.oncogenes)) + 
  labs(x="Known Oncogenes", y="MAPK activity signature") +
  theme_bw() +
  scale_colour_manual(values=palette.known.onco) +
  coord_flip()
```

7.  Add another geometric shape: points

This task consist of **adding a new layer** containing the points colored by known oncogenes. Which function should we use ? Note that there are two answers but one is more suitable for boxplot as it is adding some random noise to the positions in order to avoid overlaps.

<details>

<summary>Key words to write on your favorite browser</summary>

-   `R ggplot2 point`
-   `R ggplot2 point boxplot`

</details>

***Option 1***

```{r}
ggplot(data = clinical, aes(x=known.oncogenes, y=signature_MAPK)) +
  geom_boxplot(aes(color=known.oncogenes)) + 
  labs(x="Known Oncogenes", y="MAPK activity signature") +
  theme_bw() +
  scale_colour_manual(values=palette.known.onco) +
  coord_flip() +
  geom_point(aes(color=known.oncogenes))
```

***Option 2***

```{r}
ggplot(data = clinical, aes(x=known.oncogenes, y=signature_MAPK)) +
  geom_boxplot(aes(color=known.oncogenes)) + 
  labs(x="Known Oncogenes", y="MAPK activity signature") +
  theme_bw() +
  scale_colour_manual(values=palette.known.onco) +
  coord_flip() +
  geom_jitter(aes(color=known.oncogenes))
```

8.  Try to change the **shape** in plain triangle for all points and decrease **opacity** by half and to manage the **size** according to the tumor purity (column `Purity`).

```{r}
ggplot(data = clinical, aes(x=known.oncogenes, y=signature_MAPK)) +
  geom_boxplot(aes(color=known.oncogenes)) + 
  labs(x="Known Oncogenes", y="MAPK activity signature") +
  theme_bw() +
  scale_colour_manual(values=palette.known.onco) +
  coord_flip() +
  geom_jitter(aes(color=known.oncogenes, size=Purity), shape=17, alpha=0.5)
```

## Create a density plot

Your goal is to produce the following density plot showing the **density distributions** of the MAPK signature in function of the known oncogenes (use `palette.known.onco`). Here we introduce the customization of the plot dimensions, we want to force the plot to have be x-axis 2 times longer than the y-axis which we call the **aspect ratio**. Thanks to it, you can force to have squared plots for instance.

```{r, echo=FALSE}
ggplot(data = clinical, aes(x=signature_MAPK)) +
  geom_density(aes(color=known.oncogenes)) + 
  labs(x="MAPK activity signature") +
  theme_bw() +
  scale_colour_manual(values=palette.known.onco) +
  theme(aspect.ratio = 1/2)
```

1.  Identify what contain the axis.

<details>

<summary>Answer</summary>

The x-axis contains the value of the column `signature_MAPK` and the y-axis indicates the density of the distribution, the y-axis do not need to be specified.

</details>

2.  Make a list of the aesthetics you need to modify and layers that need to be added.

<details>

<summary>Answer</summary>

-   layers: the geometric shape, label of the x-axis, theme background, theme aspect ratio, custom color palette

-   aesthetics: x-axis , color of the lines

</details>

3.  Write your code ! (Do not forget your best friend: the Web)

```{r, eval=F}
ggplot(data = clinical, aes(x=signature_MAPK)) +
  geom_density(aes(color=known.oncogenes)) + 
  labs(x="MAPK activity signature") +
  theme_bw() +
  scale_colour_manual(values=palette.known.onco) +
  theme(aspect.ratio = 1/3)
```

## Create a bar plot

Your goal is to produce the following **bar plot** showing the number of tumor from genders (column `gender` - play with the **transparency**) harboring each known oncogene. In this case the **aspect ratio** is inverse compared to the previous one.

```{r, echo=FALSE, warning=F, message=F}
ggplot(data = clinical, aes(x=known.oncogenes)) +
  geom_bar(aes(fill=known.oncogenes, color=known.oncogenes,alpha=gender)) + 
  labs(x="Known Oncogenes") +
  theme_bw() +
  scale_fill_manual(values=palette.known.onco) +
  scale_colour_manual(values=palette.known.onco) +
  coord_flip() +
  theme(aspect.ratio = 2)
```

1.  Identify what contain the axis.

<details>

<summary>Answer</summary>

The x-axis indicates the number of sample harboring the known oncogene, the x-axis do not need to be specified. The y-axis indicates the known oncogenes. As a reminder, one would prefer to flip the coordinates but it is not mandatory.

</details>

2.  Make a list of the aesthetics you need to modify and layers that need to be added.

<details>

<summary>Answer</summary>

-   layers: the geometric shape, label of the y-axis, theme background, theme aspect ratio, custom color palette

-   aesthetics: y-axis , color of the lines, fill of the bars, transparency according to gender

</details>

3.  Write your code ! (Do not forget your best friend: the Web)

```{r, eval=F}
ggplot(data = clinical, aes(x=known.oncogenes)) +
  geom_bar(aes(fill=known.oncogenes, color=known.oncogenes,alpha=gender)) + 
  labs(x="Known Oncogenes") +
  theme_bw() +
  scale_fill_manual(values=palette.known.onco) +
  scale_colour_manual(values=palette.known.onco) +
  coord_flip() +
  theme(aspect.ratio = 2)
```

## Display expression data

The goal of this section is to manipulate the expression (`luad_expr`) and clinical datasets (`clinical`) by filtering for patients harboring specific known oncogene, having extreme MAPK activity signal. We want to observe the correlation between the signature and the genes that were used to compute it: ETV4, ETV5, DUSP6, DUSP4 and SPRY2. We will use scatter plots (basically visualizing samples with points), linear regression estimation and a heatmap using the package `corrplot`. The expression dataset was subset by filtering for genes involved from closer or farther to MAPK signalling. We will operate a clustering of the samples and genes and visualize the expression of the genes using a heatmap that we will annotate using the package `pheatmap`.

#### Prepare the data

1.  Join the clinical annotations with the expression of the MAPK signature genes.

The best way to complete this task is to i) filter for our genes of interest by declaring a variable named `gene.of.interest` containing a vector with the gene symbols and ii) **transpose** (flip columns and rows) the expression table (`luad_expr`). The function allowing this operation return a object of type `matrix` , this type is not suitable for ggplot and not convenient for easy manipulation. We need to transform it in **`data.frame`** format. iii) Create a column `sampleID` containing the sample identifiers present as **row names**. iv) Merge the 2 `data.frame`s.

Note: In R, if possible, you can transform the type of variables using functions starting with `as.` followed by the desired format. For instance you have a character string `"42"` that you want to transform in numeric, you can run `as.numeric("42")`.

i\) Create `gene.of.interest` containing the genes and filter for the rows containing the expression of these genes by creating a new variable `expression.selected`

```{r}
# Create a vector of characters using c()
gene.of.interest <- c("ETV4", "ETV5", "DUSP6", "SPRY2","DUSP4")
# The gene symbols are the row names, then the gene of interest are used as indices
expression.selected <- luad_expr[gene.of.interest,]
```

ii\) Transpose the data.frame `expression.selected`

```{r}
# Transpose the data.frame with t()
expression.selected <- t(expression.selected)
# show the type of expression.selected
class(expression.selected)
# Transform expression.selected as data.frame 
expression.selected <- as.data.frame(expression.selected)
# show the first lines of the table
head(expression.selected)
# show the type of expression.selected
class(expression.selected)
```

iii\) Create a column `sampleID` in the `data.frame` `expression.selected` containing the sample identifiers that are present in the row names

```{r}
expression.selected$sampleID <- row.names(expression.selected)
```

iv\) Merge `expression.selected` and `clinical`

```{r}
clinical <- merge(clinical,expression.selected, by="sampleID")
# Always verify the object you obtain 
head(clinical)
```

#### What are this plot ?

Analyse the following code chunks, comment all the layers of each plot, which plots would you expect ? What do you need to be able to run these codes. Ask the web or chatGPT !

***Plot 1***

```{r, eval=F}
ggscatter(clinical, x="DUSP6", y="signature_MAPK", color="known.oncogenes", size = 0.5, add="reg.line") +
  theme_light() +
  theme(legend.position = "right") +
  facet_wrap(~known.oncogenes,ncol=4) +
  stat_cor(label.y = 1) +
  scale_colour_manual(values=palette.known.onco) +
  theme(aspect.ratio = 1)
```

***Plot 2***

Comment also the lines before the plot.

```{r, echo=F}
median.signature <- median(clinical$signature_MAPK)
high.signature <- clinical$signature_MAPK >= median.signature
clinical$signature_group <- "low"
clinical$signature_group[high.signature] <- "high"
```

```{r, eval=F}
median.signature <- median(clinical$signature_MAPK)
high.signature <- clinical$signature_MAPK >= median.signature
clinical$signature_group <- "low"
clinical$signature_group[high.signature] <- "high"

ggviolin(clinical, x = "signature_group", y = "ETV4", fill = "signature_group", 
         palette = c(high=colours()[124],low=colours()[144]),
         add = "boxplot", add.params = list(fill = "white")) +
  theme_light() +
  theme(aspect.ratio = 1) +
  facet_wrap(~known.oncogenes,ncol=4, scales="free_y") +
  stat_compare_means(comparisons = list(c("high","low")), method="wilcox", label.y = 3, label="p.signif")
```

#### Correlation plot

We want to analyse the correlation between the genes of the MAPK activity signature. To this end, we need to **install** and **load** the `corrplot` package. You will use the table stored in `expression.selected`, you need to remove the column containing the sample IDs and compute the correlation between the columns and store the result in a variable.

```{r}
expression.selected$sampleID <- NULL
```

You have a very nice introduction to this package: <https://cran.r-project.org/web/packages/corrplot/vignettes/corrplot-intro.html>. All the informations are in this page but don't hesitate to search for more.

You want to obtain the following plot:

```{r, echo=F}
M <- cor(expression.selected) 
corrplot(M, addCoef.col = 'black', order = 'AOE', col = COL2('PiYG'), tl.pos = 'd', cl.pos = 'b')
```

Here is the solution:

```{r, eval=F}
# install the package that is in CRAN repository
install.packages("corrplot")
# load the library
library(corrplot)
# compute the correlation, by default it compute the correlation between columns
M <- cor(expression.selected) 
# the plot !
corrplot(M, addCoef.col = 'black', order = 'AOE', col = COL2('PiYG'), tl.pos = 'd', cl.pos = 'b')
```

#### Bonus: Heatmap

In this section, we want to visualise the gene expression across all samples using a heatmap and highlight potential expression patterns using sample and gene clustering.

<details>

<summary>What is a heatmap ?</summary>

In a heatmap, **each row** usually represents a **gene**. Each **column** represents a **sample**. The rectangles containing the values are also called "cells".

The **colour and intensity** of the cells is used to represent values of **gene expression**. So, basically, instead of numbers, we use colours. The colour will be more **intense** the lower (or higher) the value is.

To see patterns, we cluster (meaningful **reordering**) the rows and columns. This just means we **group** the samples and/or genes together based on the **similarity** of their gene expression pattern. The **dendograms** on the sides just indicate the results of clustering both genes and samples.

A clustering algorithm will group genes with a similar expression across samples together: groups of genes that are highly expressed in some samples, and lowly expressed in other samples will be clustered together. In a similar way, samples with a similar expression pattern (in general having certain sets of genes highly expressed and certain sets of genes lowly expressed) will be clustered together.

In summary, adding **clustering** to heatmaps can be useful for identifying genes that are commonly regulated, or biological signatures associated with a particular condition (for example, high MAPK activity).

</details>

```{r, echo=F}
#| fig.height: 10
## Prepare the data
# annotation of the columns that contains sample informations
# We want to show the oncogenes, the gender and the pathological stage and the 
# MAPH signature groups
# the data frame must contain row names corresponding to the sample IDs and identical to the sample IDs in the expression data.
# We create a data.frame with the relevant columns
anno.columns <- clinical[,c("known.oncogenes","pathologic_stage","gender","signature_group")]
row.names(anno.columns) <- clinical$sampleID
# For the pathological stage we want only 4 stages, we need to to remove sub categories A and B
anno.columns$pathologic_stage <- sub("[AB]$","",anno.columns$pathologic_stage)
# create a list containing the colors for each feature
colors.anno <- list(known.oncogenes=palette.known.onco,
                       pathologic_stage=c("Stage I"=colours()[404], "Stage II"=colours()[33], "Stage III"=colours()[35], "Stage IV"=colours()[24]),
                       gender=c(FEMALE=colours()[613], MALE=colours()[290]),
                       signature_group=c(high=colours()[124],low=colours()[144]))
# Create annotation for the rows containing the genes: 
# we want to indcate the genes involved in the signature
anno.rows <- data.frame(row.names=row.names(luad_expr),in.signature=ifelse(row.names(luad_expr)%in%gene.of.interest,"yes","no"))
colors.anno$in.signature <- c(yes="black",no="white")

pheatmap(luad_expr,
         cluster_rows = T, cluster_cols = T, # set to FALSE if you want to remove the dendograms
         show_colnames = F, # displaying column names
         show_rownames = T, # displaying row names
         clustering_distance_cols = 'correlation',
         clustering_distance_rows = 'correlation',
         annotation_row = anno.rows, # row (gene) annotations
         annotation_col = anno.columns, # column (sample) annotations
         annotation_colors = colors.anno, # colours for your annotations
         clustering_method = 'ward.D', # how to aggregate rows and columns depending on their distance
         color = viridis::inferno(100), # choose a colour scale for your data
         breaks = seq(-4,4,0.08),
         cutree_cols = 6, # display column clusters
         cutree_rows = 4, # display row clusters
         fontsize = 8, # change font size
         main = "Super heatmap with annotations" # a title for our heatmap
         )
```

*Step by step*

***Before starting ...***

i\) We need to find a package that allow us to draw (pretty) annotated heatmaps. Lets search on the web what we could try.

<details>

<summary>Key words to write on your favorite browser</summary>

R How to draw annotated heatmap

</details>

<details>

<summary>The package we will be using</summary>

`pheatmap`

</details>

ii\) Install the package.

```{r, eval=FALSE}
install.packages("pheatmap")
library(pheatmap)
```

***Step 1 : Prepare the data***

To be able to annotate the heatmap we need to build the annotations for the samples (`clinical`) and the genes (`genes.of.interest`) and associate colours to the categories for each feature.

i\) Annotation of the samples

Using the variable `clinical`, extract the the columns known.oncogenes, pathologic_stage, gender, signature_group in a new variable `anno.columns`.

```{r}
anno.columns <- clinical[,c("known.oncogenes","pathologic_stage","gender","signature_group")]
head(anno.columns)
```

To allow the tool to map the annotations to the samples in the expression dataset, we need to name the row with the sample identifiers that are stored in the column sampleID of `clinical`.

```{r}
row.names(anno.columns) <- clinical$sampleID
head(anno.columns)
```

To define the colours for each feature, you need to build a list using the constructor function `list()` in which each element will be a named vector of colors as you have already made to create `palette.known.onco`. For that you have to know all possible categories for each feature. `table()` is an easy and very useful function to display all categories with the number of sample per category.

`table(anno.columns$gender)`

```{r, echo = FALSE}
table(anno.columns$gender)
```

`table(anno.columns$pathologic_stage)`

```{r, echo = FALSE}
table(anno.columns$pathologic_stage)
```

etc

You can see that for the pathological stages, there are subcategories A and B that we want to regroup in order to simplify the visualisation. To group the sub-categories, we will remove the A and B from the end of the words using the function `sub()`. Try to write the code, many answers are possible.

Here are 2 possibilities:

```{r}
anno.columns$pathologic_stage <- sub("A","",anno.columns$pathologic_stage)
anno.columns$pathologic_stage <- sub("B","",anno.columns$pathologic_stage)
# The cleanest way of coding this task is
anno.columns$pathologic_stage <- sub("[AB]$","",anno.columns$pathologic_stage)
# [AB] indicates that at the first position you encounter
```

ii\) Annotations of the genes

We want to identify the genes belonging to the MAPK activity signature. To this end, we create a `data.frame` using the constructor function `data.frame()` with the **row names** corresponding to the gene names and containing one column `in.signature` containing yes or no.

Create a vector containing "yes" or "no" depending on the gene presence in the signature. Many possibilities exist but the best way is to use the function `ifelse()`. This function takes 3 arguments, an expression **comparing** the gene names (row names of the table `luad_expr`) with the genes of interest (`gene.of.interest`), what the function must return if the comparison is TRUE (the gene is in the signature, "yes"), what the function must return if the comparison is FALSE (the gene is not in the signature, "no").

To know if the elements of a vector are present in an other vector we use the operator **%in%**.

`genes.in.signature <- row.names(luad_expr) %in% gene.of.interest`

Try to write the code that allow to obtain the yes/no vector using `ifelse()` and `genes.in.signature` and keep the results in a variable called `in.signature`.

```{r}
in.signature <- ifelse(row.names(luad_expr)%in%gene.of.interest,"yes","no")
```

Now you can create the `data.frame`.

```{r}
# Create annotation for the rows containing the genes: 
# we want to indcate the genes involved in the signature
anno.rows <- data.frame(row.names=row.names(luad_expr),in.signature)
head(anno.rows)
```

Add colors for this feature to the list of colors `colors.anno`.

`colors.anno$in.signature <- c(yes="black",no="white")`

***Step 2 : Create a clustered heatmap***

Display the heatmap with default parameters (do not change any parameters) using the function `pheatmap()` and the expression object `luad_expr`.

```{r}
#| fig.height: 10
pheatmap(luad_expr)
```

By default, pheatmap applies hierarchical clustering to both rows and columns. The parameter names are quite explicit.

By looking at the help of the function (`?pheatmap`) or online, identify the parameters and their default values that:

-   Enable row or column clustering (2 parameters)

<details>

<summary>Answer</summary>

```         
cluster_cols = TRUE
cluster_rows = TRUE
```

</details>

-   specify the statistics to compute the distances between rows or columns (2 parameters)

<details>

<summary>Answer</summary>

```         
clustering_distance_rows = "euclidean"
clustering_distance_cols = "euclidean"
```

</details>

-   specify how to aggregate the rows and columns based on their distance (1 parameters)

<details>

<summary>Answer</summary>

```         
clustering_method = "complete"
```

</details>

***Step 2 : Customize your heatmap***

By analysing the default heatmap, we can notice that column labels are not readable and are not adding relevant information, we want to remove them, the row names are readable but we want to decrease the size of the font. We will test different method for clustering the rows and columns. We would like to change the colors of the cells to make them more divergent using for instance the color palette from `viridis` and we want to intensify the differences by lowering the range value amplitude (between -4 and 4), meaning that values outside of this range will display the extreme colors. We then want to annotate the samples and genes with nice colors, separate the clusters and finally add a tittle.

*i) Row and column names*

By looking at the help of the function (`?pheatmap`) or online, identify the parameters (and adapt their values) that:

-   Enable the display of the column names and disable it.

<details>

<summary>Answer</summary>

```         
show_colnames = T # default
```

</details>

-   Enable to decrease the font size

<details>

<summary>Answer</summary>

```         
fontsize = 10 # default
```

</details>

```{r}
#| fig.height: 10
pheatmap(luad_expr,
         cluster_rows = T, cluster_cols = T, # set to FALSE if you want to remove the dendograms
         show_colnames = F, # displaying column names
         show_rownames = T, # displaying row names
         fontsize = 8 # change font size
         )
```

*ii) Change the colors*

By looking at the help of the function (`?pheatmap`) or online, identify the parameters (and adapt their values) that:

-   Enable to set the color palette. In the example we used the inferno palette from the package viridis and asked for 100 colors.

<details>

<summary>Answer</summary>

```         
color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdYlBu")))(100) # Default
```

</details>

-   Enable to change the amplitude of the colors between -4 and 4

<details>

<summary>Answer</summary>

```         
breaks = NA # default
```

</details>

```{r, eval=F}
pheatmap(luad_expr,
         cluster_rows = T, cluster_cols = T, # set to FALSE if you want to remove the dendograms
         show_colnames = F, # displaying column names
         show_rownames = T, # displaying row names
         clustering_distance_cols = 'correlation',
         clustering_distance_rows = 'correlation',
         clustering_method = 'ward.D', # how to aggregate rows and columns depending on their distance
         color = viridis::inferno(100), # choose a colour scale for your data
         breaks = seq(-4,4,0.08), # seq produce a vector from -4 to 4 with a step of 0.08 (100 values)
         fontsize = 8 # change font size
         )
```

*iii) Choose the clustering methods*

Using the parameters you identified earlier for clustering methods, test several combinations to observe the impact on the results. You can find bellow the definition of the different methods. In the example, we chose to cluster following the correlation distance and the ward.D agglomeration methods.

<details>

<summary>Distance methods</summary>

**Euclidean Distance**: This is like drawing a straight line between two points. It\'s the most intuitive way to measure distance, like the shortest path "as the crow flies" between two locations on a map.

**Manhattan Distance**: Think of how you would walk in a city laid out like a grid, where you can only move horizontally or vertically. The Manhattan distance measures the total number of blocks you would walk.

**Correlation Distance**: Instead of focusing on actual distance, this measures how similar the patterns or trends are between points. It\'s useful when you care about how things change together rather than how far apart they are.

**Maximum Distance (Chebyshev)**: This only looks at the largest difference in any single direction (horizontal or vertical). Imagine trying to figure out how far two buildings are by only checking the longest side of the blocks between them.

**Canberra Distance**: This one gives more weight to differences when values are small. It's useful when small differences matter a lot in your data.

**Binary Distance**: This is used for yes/no or presence/absence type data. It simply compares if two points have the same "yes/no" values.

**Minkowski Distance**: This is a general form of both Euclidean and Manhattan distances. By changing a parameter, it can behave like either Euclidean (straight line) or Manhattan (grid-based).

</details>

<details>

<summary>Agglomeration methods</summary>

**Ward's Method (ward.D, ward.D2)**: This tries to form groups that minimize the difference within each cluster. It prefers clusters where data points are very similar to each other. It\'s great for balanced, compact groups.

**Single Linkage (single)**: This method focuses on the smallest distance between points in two clusters. It can create "chains" of points that are loosely connected, leading to long, stretched-out clusters.

**Complete Linkage (complete)**: This method uses the largest distance between points in two clusters. It prefers tight, compact clusters, and avoids the stretched-out effect of single linkage.

**Average Linkage (average/UPGMA)**: Here, we average the distances between all pairs of points in two clusters. It balances between tight and loose clusters, forming more moderate-sized groups.

**McQuitty's Method (mcquitty/WPGMA)**: Similar to average linkage, but it calculates the average in a slightly different way. It\'s another method to balance between close and distant points.

**Median Method (median/WPGMC)**: This method looks at the median distances between points, creating balanced clusters that can handle uneven data well.

**Centroid Method (centroid/UPGMC)**: This method calculates the "center" of clusters (think of the average location of all points in the cluster) and uses that to decide how to group them.

</details>

```{r, eval=F}
pheatmap(luad_expr,
         cluster_rows = T, cluster_cols = T, # set to FALSE if you want to remove the dendograms
         show_colnames = F, # displaying column names
         show_rownames = T, # displaying row names
         clustering_distance_cols = 'correlation',
         clustering_distance_rows = 'correlation',
         clustering_method = 'ward.D', # how to aggregate rows and columns depending on their distance
         fontsize = 8 # change font size
         )
```

*iv)* *Separate clusters*

By looking at the help of the function (`?pheatmap`) or online, identify the parameters (and adapt their values) that:

-   Enable to show distinct clusters, in the example, 6 sample clusters and 4 gene clusters (2 parameters)

<details>

<summary>Answer</summary>

```         
 cutree_rows = NA # Default
 cutree_cols = NA # Default
```

</details>

```{r, eval=FALSE}
pheatmap(luad_expr,
         cluster_rows = T, cluster_cols = T, # set to FALSE if you want to remove the dendograms
         show_colnames = F, # displaying column names
         show_rownames = T, # displaying row names
         clustering_distance_cols = 'correlation',
         clustering_distance_rows = 'correlation',
         annotation_row = anno.rows, # row (gene) annotations
         annotation_col = anno.columns, # column (sample) annotations
         annotation_colors = colors.anno, # colours for your annotations
         clustering_method = 'ward.D', # how to aggregate rows and columns depending on their distance
         color = viridis::inferno(100), # choose a colour scale for your data
         breaks = seq(-4,4,0.08),
         cutree_cols = 6, # display column clusters
         cutree_rows = 4, # display row clusters
         fontsize = 8 # change font size
         )
```

## Bonus: Create a KM plot

survival

ggsurvfit

## Some useful links

-   <https://ggplot2.tidyverse.org/index.html>

-   <https://ggplot2.tidyverse.org/reference/ggtheme.html>

-   <http://www.sthda.com/english/wiki/ggplot2-themes-and-background-colors-the-3-elements>

-   <https://r-graph-gallery.com/>

-   <https://www.cedricscherer.com/2019/08/05/a-ggplot2-tutorial-for-beautiful-plotting-in-r/>

-   <https://r-graph-gallery.com/ggplot2-color.html>

-   [https://larmarange.github.io/analyse-R/couleurs.html](https://larmarange.github.io/analyse-R/couleurs.html#ggsci)

-   <https://www.emilyzabor.com/tutorials/survival_analysis_in_r_tutorial.html>

-   <https://cran.r-project.org/web/packages/corrplot/vignettes/corrplot-intro.html>

-   <https://biostatsquid.com/step-by-step-heatmap-tutorial-with-pheatmap/>
